---
title: "Homework Assignment #3"
subtitle: "Exploring patterns of environmental justice"
author: "Joshua Mull"
date: last-modified
execute:
  warning: false
  message: false
format:
  html:
    toc: true
    code-fold: true
editor_options: 
  chunk_output_type: console
---


#### Load libraries 
```{r}
library(tidyverse)
library(sf)
library(stars)
library(terra)
library(tmap)
library(dplyr)
library(kableExtra)
library(patchwork)
```

#### Read in data 
```{r, quiet = TRUE}
# Use st_read to read in building data, save as variable building
building <- st_read(
  here::here("data", "gis_osm_buildings_a_free_1.gpkg"),
  query = "SELECT * FROM gis_osm_buildings_a_free_1
           WHERE (type IS NULL AND name IS NULL) 
           OR type IN ('residential', 'apartments', 'house', 'static_caravan', 'detached')"
)

# Use st_read to read in road data, save as variable roads
roads <-st_read(here::here("data", "gis_osm_roads_free_1.gpkg"), query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'")

# Use rast from the terra package to read in tile h08v05 date 02/07/21, save as variable tile_8_5_07
tile_8_5_07 <- read_stars(here::here("data", "VNP46A1", "VNP46A1.A2021038.h08v05.001.2021039064328.tif"))


# Use rast from the terra package to read in tile h08v06 date 02/07/21, save as variable tile_8_6_07
tile_8_6_07 <- read_stars(here::here("data", "VNP46A1", "VNP46A1.A2021038.h08v06.001.2021039064329.tif"))

# Use rast from the terra package to read in tile h08v05 date 02/16/21, save as variable tile_8_5_16
tile_8_5_16 <- read_stars(here::here("data", "VNP46A1", "VNP46A1.A2021047.h08v05.001.2021048091106.tif"))

# Use rast from the terra package to read in tile h08v06 date 02/16/21, save as variable tile_8_6_16
tile_8_6_16 <- read_stars(here::here("data", "VNP46A1", "VNP46A1.A2021047.h08v06.001.2021048091105.tif"))

# Use st_read to read in income data
income <- st_read(here::here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X19_INCOME")

# Use st_read to read in road data, save as variable 
texas_cens <- st_read(here::here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "ACS_2019_5YR_TRACT_48_TEXAS")

# Read in ejscreen data for base-map 
ejscreen <- st_read(here::here("data", "ejscreen", "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))

```

```{r, warning=TRUE}
roads <- st_transform(roads, crs = 3083)
building <- st_transform(building, crs = 3083)
ejscreen <- st_transform(ejscreen, crs = 3083)
texas_cens <- st_transform(texas_cens, crs = 3083)
```

#### Create background map 
```{r}
# Use filter to filter for counties around houstin, this is for my base-map
houston <- ejscreen %>% 
  filter((STATE_NAME == "Texas") & (CNTY_NAME %in% c("Harris County", "Liberty County", "Austin County", "Waller County", "Montgomery County", "Chambers County", "Fort Bend County", "Galveston County", "Brazoria County", "Washington County", "Grimes County", "Burleson County", "Brazos County", "San Jacinto County", "Austin County", "Colorado County", "Matagorda County", "Wharton County", "Jackson County", "Polk County", "Walker County", "Hardin County", "Jefferson County", "Tyler County")))

```


#### Combine tiles for each day, spatially subset the blackout mask to the Houston area as defined by the following coordinates:
(-96.5, 29), (-96.5, 30.5), (-94.5, 30.5), (-94.5, 29)
```{r, fig.width=10, fig.height=8}
# Merge tiles to make one frame for each day, the 7th and the 16th
tile_merge_07 <- st_mosaic(tile_8_5_07, tile_8_6_07)
tile_merge_16 <- st_mosaic(tile_8_5_16, tile_8_6_16)

# Define the desired bounding box for Houston
houston_bbox <- st_bbox(c(xmin = -96.5, xmax = -94.5, ymin = 29, ymax = 30.5), crs = st_crs(tile_merge_07))

# Use st_as_sf
houston_sf <- st_as_sfc(houston_bbox) %>%
  st_transform(st_crs(tile_merge_07))

# Crop the raster using the bounding box
tile_merge_07 <- st_crop(tile_merge_07, houston_bbox)
tile_merge_16 <- st_crop(tile_merge_16, houston_bbox)

# make a copy of each and save them to new variables. I want to use the copy to make my first two maps. In order to see the base-map, I have to element some lower values, around 50 nW cm² sr⁻¹. If not, my base-map will be completely covered
tile_merge_07_copy <- tile_merge_07
tile_merge_16_copy <- tile_merge_16 

# Set values less than 200 nW cm² sr⁻¹ to NA for mapping on top of ejscreen data. This excludes the backgroud from covering the ejscreen data, save as new variable. Note: I am not sure of the minimum light detected for the background. This number came from trial and error
tile_merge_07_copy[tile_merge_07_copy <= 50] <- NA
tile_merge_16_copy[tile_merge_16_copy <= 50] <- NA

# Plot both days to compare. This helps understand what the difference mask should look like 
hou_map_07 <- tm_shape(houston, 
         bbox = houston_bbox) + 
  tm_polygons(fill = "#000731",
              col = "#707483",
              lwd = .3) +
tm_shape(tile_merge_07_copy) + 
  tm_raster(palette = c("#bea878", "#fffbf4"),
            breaks =c(51, 100, 200, 300, 500, 1000, 2000, 4000, 6000, 8000, Inf)) + 
  tm_layout(bg.color = "#010224",
            legend.show = FALSE)

hou_map_16 <- tm_shape(houston, 
         bbox = houston_bbox) + 
  tm_polygons(fill = "#000731",
              col = "#707483",
              lwd = .3) +
tm_shape(tile_merge_16_copy) + 
  tm_raster(palette = c("#bea878", "#fffbf4"),
            breaks =c(51, 100, 200, 300, 500, 1000, 2000, 4000, 6000, 8000, Inf)) + 
  tm_layout(bg.color = "#010224",
            legend.show = FALSE)

tmap_arrange(hou_map_07, hou_map_16, ncol = 2, asp = 1)
```

#### find the change in night lights intensity (presumably) caused by the storm
```{r, fig.width=10, fig.height=8}
# Subtract tile day 16 from tile day 7 to created difference raster mask 
diff_rast_mask <- (tile_merge_07 - tile_merge_16)
```

#### Reclassify the difference raster, assuming that any location that experienced a drop of more than 200 nW cm-2sr-1 experienced a blackout,
assign NA to all locations that experienced a drop of less than 200 nW cm-2sr-1 change, vectorize the blackout mask
```{r, fig.width=10, fig.height=8}
# Assign NA to locations with a drop of less than 200 nW cm² sr⁻¹ 
diff_rast_mask[diff_rast_mask <= 200] <- NA 

# Vectorize blackout mask with as.polygons and st_as_sf to make it vector data instead of raster
diff_poly_mask <- as.polygons(rast(diff_rast_mask)) %>% 
  st_as_sf() %>% 
  st_transform(crs = 3083)
```
 
#### Re-project the cropped blackout dataset to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area)


#### Plot Blackout Mask
```{r, fig.width=10, fig.height=8}
# Map the blackout data
blackout_map <- tm_shape(houston, 
         bbox = diff_rast_mask) + 
  tm_polygons(fill = "seashell1",
              col = "grey80") + 
  tm_shape(diff_poly_mask) +
  tm_polygons(fill = "brown", 
              alpha = .7,
              border.col = NA) + 
  tm_layout(bg.color = "#E0FFFF") + 
  tm_title(text = "Blackout Areas Houston, TX",
           fontfamily = "fantasy",
           fontface = "bold",
           position = c("left", "top"))

print(blackout_map)
```

## Exclude highways from the cropped blackout mask

#### Identify areas within 200m of all highways
```{r}
# Use st_buffer to make road buffer of 200m, unionize the buffer, transform to 3083 so it matches the mask, make all polygons valid, make simple feature 
roads_200 <- st_buffer(roads, dist = 200) %>% 
  st_union() %>% 
  st_transform(crs = 3083) %>% 
  st_make_valid() %>% 
  st_as_sf()

# Use st_difference to eliminate roads from the mask
black_map_no_roads <- st_difference(diff_poly_mask, roads_200)

# Plot to check 
map_no_roads <- tm_shape(black_map_no_roads) +
  tm_polygons()


```


#### Identify homes likely impacted by blackouts

```{r}
# Use st_intersects to see which homes were affected by the black out 
blkout_home <- st_intersects(black_map_no_roads, building)

# Unlist the list of homes and keep only unique values, this way no home is in multiple polygons. That probably would not be the case but I wanted to make sure. Unlist creates a list of index values from the building data frame 
blkout_index <- blkout_home %>%
  unlist() %>%
  unique()

# use the index to filter only homes that experienced a blackout
affected_buildings <- building[blkout_index, ]

# Map the blackout data
blackout_homes <- tm_shape(houston, 
         bbox = houston_bbox) + 
  tm_polygons(fill = "white",
              col = "grey80", 
              lwd = .5) + 
  tm_shape(affected_buildings) +
  tm_polygons(fill = "brown", 
              alpha = 1,
              border.col = NA) + 
  tm_layout(bg.color = "#E0FFFF", 
            legend.position = c("right", "bottom")) + 
  tm_title(text = "Blackout Homes Houston, TX",
           fontfamily = "fantasy",
           fontface = "bold",
           position = c("left", "top"),
           size =7,
           bg.color = "white") + 
  tm_scalebar(position = c("left", "bottom")) + 
  tm_add_legend(labels = "Blackout Homes",
                type = "fill", 
                col = "brown") + 
  tm_compass(position = c(.9, .1))
```

#### Estimate the number of homes affected 
```{r}
# Find total of homes affected
homes_total <- nrow(affected_buildings)

# Create a summary data frame 
summary_table_data <- data.frame( Description = "Total Affected Homes", Total = homes_total)

summary_table <- summary_table_data %>% kbl(col.names = c("Description", "Total"), digits = 2) %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

print(summary_table)
```


#### Select Median Income and Add It To Census Tracts
```{r}
median_income <- income %>%
  select(GEOID, B19013e1)

# Join income with the geo layer
income_geo <- left_join(texas_cens, median_income, join_by(GEOID_Data == GEOID))

income_geo <- st_transform(income_geo, crs = 3083)
```

#### Identify Census Tracts That Experienced Blackouts
```{r}
census_blkout <- st_intersects(black_map_no_roads, income_geo)

census_index <- census_blkout %>% 
  unlist() %>%
  unique()

tracts_affect <- income_geo[census_index,]

census_map <- tm_shape(tracts_affect) + 
  tm_polygons(fill = "B19013e1",
              palette = c("lightyellow", "darkgreen"))

census_map
```

#### Plot of median household incomes for census tracts affected and not affected
```{r}
# Summary income of homes affected 
home_inc_aff <- ggplot(tracts_affect, aes(x = tracts_affect$B19013e1)) + geom_histogram()

# View 
home_inc_aff

# Find the number of homes tracts not affected 
tracts_no_affect <- income_geo %>% 
  filter(!GEOID %in% tracts_affect$GEOID)

# Summary income of homes affected 
home_inc_no_aff <- ggplot(tracts_no_affect, aes(x = tracts_no_affect$B19013e1)) + geom_histogram()

# Plot 
home_inc_aff + home_inc_no_aff

summary(tracts_affect$B19013e1)

summary(tracts_no_affect$B19013e1)
```

